---
title: "cm009 Exercises: tidy data"
output: html_notebook
editor_options: 
  chunk_output_type: md_document
---

getwd() gets working directory
setwd("/Users/ernest91/Desktop/SPPH545/Class Notes") 
    Sets working directory to 'Class Notes'

```{r}
suppressPackageStartupMessages(library(tidyverse))
```

## Reading and Writing Data: Exercises

Make a tibble of letters, their order in the alphabet, and then a pasting of the two columns together.

```{r}
tibble(let = letters,
       num = 1:length(letters),
       comb = paste(let, num))

tibble(let = letters,
       num = 1:length(letters),
       comb = paste0(let, num)) # no space between let and num
```

Make a tibble of three names and commute times.

```{r}
tribble(
    ~name, ~time,
    "Frank", 30,
    "Lisa", 15,
    "Fred", 40
) # ~ indicates var names we will make

```


Write the `iris` data frame as a `csv`. 

```{r}
write_csv(iris, "iris.csv")
```

Write the `iris` data frame to a file delimited by a dollar sign. 

```{r}
write_delim(iris, "iris.txt", delim = "$") # separates each column with a $ sign
```

Read the dollar-delimited `iris` data to a tibble.

```{r}
read_delim("iris.txt", delim = "$")
```

Read these three LOTR csv's, saving them to `lotr1`, `lotr2`, and `lotr3`:

- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv

```{r}
lotr1 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")

lotr2 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")

lotr3 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv") 
```

## `gather()`

(Exercises largely based off of Jenny Bryan's [gather tutorial](https://github.com/jennybc/lotr-tidy/blob/master/02-gather.md))

This function is useful for making untidy data tidy (so that computers can more easily crunch the numbers).

1. Combine the three LOTR untidy tables (`lotr1`, `lotr2`, `lotr3`) to a single untidy table by stacking them.  

```{r}
# Before binding and stacking the data frames, column names have to be same - dimensions, variable type, etc.

lotr_untidy <- bind_rows(lotr1, lotr2, lotr3)

```

2. Convert to tidy. Also try this by specifying columns as a range, and with the `contains()` function.

```{r}
gather(lotr_untidy, key = "Gender", # our new categorical variable
                    value="Word", # names the column  (count)
                    Female, Male # variables that make up the categorical variable
)

gather(lotr_untidy, key = "Gender", # our new categorical variable
                    value="Word", # names the column  (words counts)
                    Female:Male # variables that make up the categorical variable
)

gather(lotr_untidy, key = "Gender", # our new categorical variable
                    value="Word", # names the column  (word counts)
                    contains("ale") # looks for words that contain "ale"
)

```

3. Try again (bind and tidy the three untidy data frames), but without knowing how many tables there are originally. 
    - The additional work here does not require any additional tools from the tidyverse, but instead uses a `do.call` from base R -- a useful tool in data analysis when the number of "items" is variable/unknown, or quite large. 

```{r}
lotr_list <- list(lotr1, lotr2, lotr3)
lotr_list

do.call(bind_rows, lotr_list) # useful if we dont know how many components of a function we are adding together
```

## `spread()`

(Exercises largely based off of Jenny Bryan's [spread tutorial](https://github.com/jennybc/lotr-tidy/blob/master/03-spread.md))

This function is useful for making tidy data untidy (to be more pleasing to the eye).

Read in the tidy LOTR data (despite having just made it):

```{r}
lotr_tidy <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/lotr_tidy.csv")
```

Get word counts across "Race". Then try "Gender". 

```{r}
spread(lotr_tidy, key = "Race", value = "Words") # Takes all race, spreads across in columns and give word counts for each

```

Now try combining race and gender. Use `unite()` from `tidyr` instead of `paste()`. 

```{r} 
# combine race and gender (specify these columns we want to unite) and make new column called race_gender

lotr_tidy %>%
    unite(Race_Gender, Race, Gender) %>%
    spread(key = "Race-Gender", value = "Words")

lotr_tidy %>%
    mutate(x=rnorm(nrow(lotr_tidy))) %>%
    spread(key = "Gender", value="x")

```

## Other `tidyr` goodies

Check out the Examples in the documentation to explore the following.

`expand` vs `complete` (trim vs keep everything). Together with `nesting`. Check out the Examples in the `expand` documentation.

```{r}
expand(mtcars, vs, cyl) 
# returns all possible combinations of vs and cyl, even if not in dataframe (ie. vs = 1 and cyl =8 does not exist)

df <- tibble(
  year   = c(2010, 2010, 2010, 2010, 2012, 2012, 2012),
  qtr    = c(   1,    2,    3,    4,    1,    2,    3),
  return = rnorm(7)
)

df

df %>% expand(year, qtr)
df %>% expand(year = 2010:2012, qtr)
df %>% expand(year = full_seq(year, 1), qtr)
df %>% complete(year = full_seq(year, 1), qtr)

```

`separate_rows`: useful when you have a variable number of entries in a "cell".

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "d,e,f", "g,h"),
  z = c("1", "2,3,4", "5,6"),
  stringsAsFactors = FALSE
)
separate_rows(df, y, z, convert = TRUE)
```

`unite` and `separate`.

```{r}
library(dplyr)
unite_(mtcars, "vs_am", c("vs","am"))

# Separate is the complement of unite
mtcars %>%
  unite(vs_am, vs, am) %>%
  separate(vs_am, c("vs", "am"))



```

`uncount` (as the opposite of `dplyr::count()`)

```{r}

```


`drop_na` and `replace_na`

```{r}

```

`fill`

```{r}
df <- data.frame(Month = 1:12, Year = c(2000, rep(NA, 11)))
df %>% fill(Year)
```


`full_seq`

```{r}

```


## Time remaining?

Time permitting, do [this exercise](https://github.com/jennybc/lotr-tidy/blob/master/02-gather.md#exercises) to practice tidying data. 